---
title: "LHS610 FINAL PROJECT"
author: "Vardh Jain"
date: "2023-04-20"
output: html_document
---

```{r}
library(tidyverse)
library(ggplot2)
```

```{r}
dataset = read_csv('FINAL-heart_disease_uci.csv')
```
```{r}
dataset = dataset %>% 
  mutate(has_heart_disease = if_else(num == 0, 'No', 'Yes'))
```

Unstratified Analysis

What is the relationship between the exposure variable and the outcome variable?

Generate a visualization in ggplot to evaluate this relationship.
Run an appropriate statistical test to evaluate this relationship.
Provide an interpretation for both the visualization and the statistical test.

```{r}
dataset %>%
  filter(!is.na(ca)) %>%  
  ggplot() +
    aes(x = ca, fill = has_heart_disease) +
    geom_bar(position = "dodge") +
    labs(x = "Number of Major Vessels colored by Fluoroscopy (0-3)",
         y = "Number of Patients",
         fill = "Heart Disease",
         title = "Relationship between Number of Major Vessels colored\nby Fluoroscopy and Presence of Heart Disease") +
    theme_linedraw()

chisq.test(table(dataset$ca, dataset$has_heart_disease))
```

Interpretation: 
There are a significantly higher number of patients without heart disease as compared to those with heart disease when there are no major vessels colored by fluoroscopy. However, as the number of major vessels colored by fluroscopy continue to increase from 1-3, there are consistently more patients with heart disease.

The null hypothesis is that there is no relationship between the number of major vessels colored by fluoroscopy and the presence of heart disease. The p-value signifies that there is a extremely low chance of 8.806e-16 (which is very close to zero) that we would observe this data (or data more extreme than what we found) based on random chance alone. Therefore, it can be concluded that the result is statistically significant.





Stratified Analysis (Confounder 1):

Repeat the steps in Unstratified Analysis to evaluate the relationship between the exposure and the outcome stratified by one confounder.
For the stratified analysis, you will be expected to “facet” by your confounding variable to evaluate the relationship between the exposure and the outcome in each subgroup.
If your confounding variable is continuous, then you should divide this confounding variable into groups (e.g., decades of age or quintiles of age).
There is no need to run a statistical test. The appropriate test would involve regression, which has not been covered.

```{r}
bin_labels = c('20-30', '30-40', '40-50', '50-60', '60-70', '70-80')
age_decades = cut(dataset$age, breaks = seq(20, 80, by = 10), labels = bin_labels,
                  right = FALSE, include.lowest = TRUE)

dataset %>%
  #mutate(age_categories = case_when(age <= 39 ~ 'Adults',
                   #age <= 59 ~ 'Middle Age',
                   #TRUE ~ 'Old Age')) %>% 
  mutate(age_decades) %>% 
  filter(!is.na(ca)) %>%  
  ggplot() +
    aes(x = factor(ca), fill = factor(has_heart_disease)) +
    geom_bar(position = "dodge") +
    facet_wrap(~age_decades) +
    labs(x = "Number of Major Vessels colored by Fluoroscopy (0-3)",
         y = "Number of Patients",
         fill = "Heart Disease",
         title = "Relationship between Number of Major Vessels colored by Fluoroscopy\nand Presence of Heart Disease stratified by Age in years") +
    theme_linedraw()
```

Interpretation:
Stratifying the relationship by age gives us the picture that the most number of patients with heart disease are found to be in age groups of 50-60 and 60-70 years. This can mean that age does influence the risk the patient has of developing heart disease. In these high risk age groups, a relationship does still exist between the number of major vessels colored by fluroscopy and the presence of heart disease with more patients having heart disease as the vessels affected increase.


Stratified Analysis (Confounder 2):

Repeat the steps in Unstratified Analysis to evaluate the relationship between the exposure and the outcome stratified by another confounder.
For the stratified analysis, you will be expected to “facet” by your confounding variable to evaluate the relationship between the exposure and the outcome in each subgroup.
If your confounding variable is continuous, then you should divide this confounding variable into groups (e.g., decades of age or quintiles of age).
There is no need to run a statistical test. The appropriate test would involve regression, which has not been covered.

```{r}
dataset %>%
  filter(!is.na(ca)) %>%  
  ggplot() +
    aes(x = factor(ca), fill = factor(has_heart_disease)) +
    geom_bar(position = "dodge") +
    facet_wrap(~sex) +
    labs(x = "Number of Major Vessels Colored by Fluoroscopy (0-3)",
         y = "Number of Patients",
         fill = "Heart Disease Presence",
         title = "Relationship between Number of Major Vessels colored by\nFluoroscopy and Presence of Heart Disease stratified by Sex") +
    theme_linedraw()
```

Interpretation:
Stratifying the relationship by sex gives us a clear interpretation that there are more number of males that with heart diseases than females. This can mean that sex does influence the risk the patient has of developing heart disease. In these sex groups, the relationship continues to exist between the number of major vessels colored by fluroscopy and the presence of heart disease with more patients having heart disease as the vessels affected increase.

